---
title: "Consumer Pricing Analysis for 2024: St. Louis Fed API"
author: "Marc Fridson"
format: html
editor: source
---

```{r}
library(httr2)
library(tidyverse)
library(lubridate)
```

Using the St. Louis Fed API, I will collect CPI (Consumer Price Index) data for 2024.

Here I define the API key and the request.

```{r}
fred_api_key <- "7af81fb6a64a50bcc9f1ffe52402d8d4"
fred_base_url <- "https://api.stlouisfed.org/fred/series/observations"

series_id <- "CPIAUCSL"  # Series ID for Consumer Price Index for All Urban Consumers

fred_req <- fred_base_url |>
  request() |>
  req_url_query(
    api_key = fred_api_key,
    series_id = series_id,
    file_type = "json",
    observation_start = "2024-01-01",  # Starting date
    observation_end = "2024-11-03"       # Ending date
  )
```

Execute the defined request and structure into data frame.

```{r}

fred_resp <- fred_req |> req_perform() |> resp_body_json()


cpi_data <- fred_resp$observations |> tibblify()


cpi_data <- cpi_data |>
  mutate(date = as_date(date),
         value = as.numeric(value))

cpi_data
```

Creating a graph of CPI values and trends from 2024 to date.

```{r}

cpi_data |>
  ggplot(aes(x = date, y = value, group = 1)) +
  geom_line()
```

**The CPI for 2024 has steadily increased, increasing the most from January through March. In April and May it was pretty flat, but has increased from June through now (11/3).**
