---
title: "NWS Code Vignette"
author: "Sandra Dela Cruz"
format: html
---

**Introduction**

The National Weather Service (NWS) API offers the public a wide range of essential weather data nationwide. This API contains data like date, time, temperature, rain precipitation, and types of weather condition.

The documentation of this API can be found here:

<https://weather-gov.github.io/api/general-faqs>

In this vignette we will explore the 7 day forecast of Santa Catalina Island, located in California, extract the data and make a plot out of it using the `tidyverse` and `httr2` packages.

```{r}
#Let's start by loading our package and defining some code 
library(tidyverse)
library(httr2)

#create base url
nws_base_url = "https://api.weather.gov"
```

We extracted our coordinates for Santa Catalina Island from the NWS as well using their search button, which can be accessed here: <https://www.weather.gov/>

```{r}
#now let's create a request
catalina_response = request(nws_base_url) |>
  req_url_path_append(
    'points',
    '33.40421,-118.41456'
  ) |>
  req_perform()
```

Let's now explore our lists to find out where to extract our data for our daily weather forecast. If you'll notice, we have first a list of 5 which contains `context`, `id`, `type`, `geometry`, and `properties`. But we'll focus on the `properties` list since this is where we'll find our information for our daily weather forecast.

```{r}
#a glimpse of our list
catalina_response |>
  resp_body_json() |> #to extract body of json
  glimpse()
```

Now let's extract from `properties` the list of `forecast`. Perform the request and fetch the response back to the server, like we did before, and have a look at our list again. Now we have a list of 4 and notice that we have `periods` under `properties` which has a list of 14 in it.

```{r}
catalina_url = catalina_response |>
  resp_body_json() |>
  pluck('properties', 'forecast')

catalina_forecast_response = request(catalina_url) |>
  req_perform()

catalina_forecast_response |>
  resp_body_json() |>
  glimpse() #to view list
```

Extract the `periods` list from `properties` and let's take a look at what are nested inside this list for us to create a table with the information we want to answer.

```{r}
catalina_7day = catalina_forecast_response |>
  resp_body_json() |>
  pluck('properties', 'periods') |>
  glimpse()
```

Now that we have seen what's nested inside, let's create a data frame which will contain our 7 day forecast.

```{r}
#create data frame for 7 day forecast
catalina_7day_night = catalina_forecast_response |>
  resp_body_json() |>
  pluck('properties', 'periods') |>
  map_dfr(
    \(x) {
      tibble(
        day = x |> pluck('name'),
        start_time = x |> pluck('startTime'),
        end_time = x |> pluck('endTime'),
        temp_F = x |> pluck('temperature'),
        type = x |> pluck('shortForecast')
      )
    }
  )|>
  mutate(
    start_time = ymd_hms(start_time),
    end_time = ymd_hms(end_time)
  )
```

Notice that our data frame consist of weathers in both day and night. Now we'll separate them into 2 data frames, so we'll know and see the pattern of our temperature in both.

```{r}
catalina_7day = catalina_7day_night [seq(1, nrow(catalina_7day_night), by = 2), ]

catalina_7night = catalina_7day_night [seq(2, nrow(catalina_7day_night), by = 2), ]

#create plot for day temp
ggplot(catalina_7day, aes(fct_inorder(day), temp_F)) +
  geom_line(group = 1) +
  geom_point() +
  labs(title = "7 day Weather Forecast of Santa Catalina Island, CA",
       subtitle = "Days Forecast",
       x = "",
       y = "Temperature in F") +
  theme_minimal() + 
  scale_y_continuous(breaks = seq(0, 100, by = 1))

#create plot for night temp
ggplot(catalina_7night, aes(fct_inorder(day), temp_F)) +
  geom_line(group = 1) +
  geom_point() +
  labs(title = "7 day Weather Forecast of Santa Catalina Island, CA",
       subtitle = "Night Forecast",
       x = "",
       y = "Temperature in F") +
  theme_minimal() + 
  scale_y_continuous(breaks = seq(0, 100, by = 1))
```

**Conclusion**

We can conclude with these plots that the temperature suddenly drops down around 40 - 50's once night time sets in Santa Catalina Island. It feels like a breath of fresh air after the scorching sun rises in that island.
